---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyr)
library(dplyr)
library(here)
library(data.table)
library(splitstackshape)
#remotes::install_github("CDK-R/rinchi")
library(rinchi)

#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("ChemmineR")
library(ChemmineR)
library(chemodiv)

```

Load in informatino from MN

clusterinfo contains the summary of all information:
```{r}

feature_table <- read.csv(here("data","xcms_all_msms_matched_only.tsv"),sep = "\t")
Feature_Info <- data.frame(FT_ID=feature_table$Row.names,num_FT_ID =as.numeric(gsub("FT","",feature_table$Row.names)))

nrow(Feature_Info)

GNPS_files <- list.files(paste0(here(),"/data/gnps_output/"),recursive =T,full.names = T)
clusterinfo <- read.csv(GNPS_files[grepl("cluster",GNPS_files)],sep = "\t")
gnps_edges <- read.csv(GNPS_files[grepl("networkedges_selfloop",GNPS_files)],sep = "\t")

Feature_Info <- left_join(Feature_Info,clusterinfo[,c("cluster.index","parent.mass","precursor.charge","precursor.mass","RTConsensus","RTMean","RTStdErr","Correlated.Features.Group.ID","componentindex","LibraryID","MQScore","Smiles")],join_by(num_FT_ID == cluster.index))

Feature_Info
names(Feature_Info)
```

Load in Info from Sirius

```{r}
#Read in Sirius Formulas Data - top hit 
sirius_formulas <- read.csv(here("data","sirius_output","formula_identifications.tsv"),sep= "\t")

Feature_Info_form <- left_join(Feature_Info,sirius_formulas %>% group_by(featureId) %>% summarise(Unique_Formulas = n_distinct(molecularFormula)),join_by(FT_ID==featureId))


#Read in the in silico hits made by Sirius
sirius_structures <- read.csv(here("data","sirius_output","compound_identifications.tsv"),sep= "\t")

Feature_Info_form_struct <- left_join(Feature_Info_form,sirius_structures %>% group_by(featureId) %>% summarise(Unique_Smiles = n_distinct(smiles)),join_by(FT_ID==featureId))

#read in Canopus_predictures as NPCassyfire classes
canopus_predictions <- read.csv(here("data","sirius_output","canopus_compound_summary.tsv"),sep= "\t")

Feature_Info_form_struct_path <- left_join(Feature_Info_form_struct,canopus_predictions %>% group_by(featureId) %>% summarise(Unique_NPClasses = n_distinct(NPC.pathway)),join_by(FT_ID==featureId))

#Read in sirius predicted fragments
sirius_pred <- read.csv(here("data","sirius_output","predictions_neg.tsv"),sep= "\t")

#add in FT ID info 
sirius_pred <- left_join(sirius_pred,sirius_formulas %>% select(featureId,id),join_by(id==id))
sirius_pred$Pred_Frag <- as.numeric(!sirius_pred$ClassyFire.Organic.compounds=="N/A")

Feature_Info_form_struct_path_frag <- left_join(Feature_Info_form_struct_path,sirius_pred %>% filter(Pred_Frag == 1) %>% group_by(featureId) %>% summarise(Unique_fragments = sum(Pred_Frag)),join_by(FT_ID==featureId))

setDT(Feature_Info_form_struct_path_frag)

Feature_Info_form_struct_path_frag
```


```{r}

#90 features got library hits

table(!Feature_Info_form_struct_path_frag$Smiles == "N/A")

#221 recieved Sirius Predictions
table(Feature_Info_form_struct_path_frag$Unique_Smiles>0)

#327 recieved class predictions
table(Feature_Info_form_struct_path_frag$Unique_NPClasses>0)

#221  recieved 326 predictions
table(Feature_Info_form_struct_path_frag$Unique_fragments >0)

```

Now to summarize the information into a single compound x compound matrix:

1) Camera Similarity - Basically the compound group identity matrix 
2) GNPS Node by Node MS matrix - easily downloaded
3) Structure x structure Similarity based the smiles output from both GNPS and Sirius in silico predictions
4) Pathway by Pathway Similarity this has to be done at different levels:
  a) Pathway - Feature x 7 pathways (fatty acids,polyketides,shikimates–phenylpropanoids,terpenoids,alkaloids,amino acids/peptides,carbohydrates)
  b) Superclass - Feature x 70
  c) Class - Feature by 672 Classes

Total Similarity will be an average of each of these 4 classes
  

*** Note it is possible to propigate uncertainty:
  For each matrix we can include all instances of a feature - multiple scans within a GNPS network
  For siriius each instance of a Feature + the value for x top hits 
  Matricies are ultimately combined for each replicate to get a mean compound by compound similarity.
  

1) MS1 Compound Group similarity - this groups compounds together if they are in the same CAMERA grouping, that is if they coellute with the same RT and if they having correlated intensities across scans.

```{r}
Camera_Matrix <- Feature_Info %>% select("FT_ID","Correlated.Features.Group.ID") %>% mutate(Present=1) %>% pivot_wider(id_cols = FT_ID,names_from = "Correlated.Features.Group.ID",names_prefix = "Camera_Group_",values_from = "Present",values_fill = 0)

Camera_Matrix
row.names(Camera_Matrix) <- Camera_Matrix$FT_ID

#get list of all pairwise comparisons
pairwise_comps <- apply(combn(Feature_Info$FT_ID,2),2,paste,collapse='_')

pairwise_comps_df <- cSplit(data.frame(combined=pairwise_comps),"combined",sep= "_",drop = T)
pairwise_comps_df <- rbind(pairwise_comps_df,data.frame(combined_1 = Feature_Info$FT_ID,combined_2 = Feature_Info$FT_ID))

pairwise_comps_df_all <- unique(rbind(pairwise_comps_df,data.frame(combined_1 = pairwise_comps_df$combined_2,combined_2 = pairwise_comps_df$combined_1)))

#Merge on Camera groupings
pairwise_comps_df_all_camera <- left_join(pairwise_comps_df_all,Feature_Info %>% select(FT_ID,Correlated.Features.Group.ID) %>% rename(combined_1=FT_ID,PCgroup_1=Correlated.Features.Group.ID),by = join_by(combined_1))

pairwise_comps_df_all_camera <- left_join(pairwise_comps_df_all_camera,Feature_Info %>% select(FT_ID,Correlated.Features.Group.ID) %>% rename(combined_2=FT_ID,PCgroup_2=Correlated.Features.Group.ID),by = join_by(combined_2))

pairwise_comps_df_all_camera$Camera_Similarity <- 0
pairwise_comps_df_all_camera$Camera_Similarity[pairwise_comps_df_all_camera$PCgroup_1 == pairwise_comps_df_all_camera$PCgroup_2] <- 1

pairwise_comps_df_all_camera <- pairwise_comps_df_all_camera %>% select(combined_1,combined_2,Camera_Similarity) %>% rename(FT_ID_1=combined_1,FT_ID_2=combined_2)

pairwise_comps_df_all_camera
```


2) GNPS Node by Node MS matrix - This is the Cosine spectral similarity between all nodes - Note most comparisons are NA
```{r}
gnps_edges <- read.csv(GNPS_files[grepl("networkedges_selfloop",GNPS_files)],sep = "\t")

table(Feature_Info$num_FT_ID %in% gnps_edges$CLUSTERID1)
table(Feature_Info$num_FT_ID %in% gnps_edges$CLUSTERID2)

gnps_edges_simplified <- left_join(
    left_join(gnps_edges,Feature_Info %>% mutate(CLUSTERID1_FT_ID = FT_ID) %>% select("num_FT_ID","CLUSTERID1_FT_ID"),join_by(CLUSTERID1==num_FT_ID),keep = T),
    Feature_Info %>% mutate(CLUSTERID2_FT_ID = FT_ID) %>% select("num_FT_ID","CLUSTERID2_FT_ID"),join_by(CLUSTERID2==num_FT_ID),keep = T) %>%
  select(CLUSTERID1_FT_ID,CLUSTERID2_FT_ID,Cosine) 

gnps_edges_simplified_self_loop <- unique(rbind(gnps_edges_simplified,data.frame(CLUSTERID1_FT_ID = Feature_Info$FT_ID,CLUSTERID2_FT_ID = Feature_Info$FT_ID,Cosine=1)))


pairwise_comps_df_all_camera_cosine <- left_join(pairwise_comps_df_all_camera,gnps_edges_simplified_self_loop,join_by(FT_ID_1==CLUSTERID1_FT_ID,FT_ID_2==CLUSTERID2_FT_ID))


pairwise_comps_df_all_camera_cosine

# Cosine data for 2020 compound x compound comparisons. Rest are 0, highly zero inflated matrix!
table(is.na(pairwise_comps_df_all_camera_cosine$Cosine))

```
3) Library and Insilico Similarity 
    get a list of unique strucutres
    Use ChemoDiv package to calculate a comp distance based on "PubChemFingerprints"
    

```{r}
gnps_hits <- Feature_Info %>% select("FT_ID","Smiles") %>% filter(Smiles != "N/A") %>% rename(smiles=Smiles)

sirius_hits <- sirius_structures %>% rename(FT_ID=featureId) %>% select("FT_ID","smiles")

#get a list of unique smile hits
unique_structures <- unique(rbind(gnps_hits,sirius_hits))
unique_structures$compound <- paste0("compound_",1:nrow(unique_structures))
unique_structures <- unique_structures[unique_structures$smiles !=" ",]
unique_structures <- unique_structures %>% filter(smiles != " ",smiles != "c")

#convert smiles to inchi-key
unique_structures <- cbind(unique_structures,data.frame(inchikey=sapply(unique_structures$smiles, get.inchi.key),inchi=sapply(unique_structures$smiles, get.inchi)))
#use compDis to determine calculate the differences between smiles based on the PubChemFingerprints

unique_structuresDis <- compDis(compoundData = unique_structures[,c("compound","smiles","inchikey")],
                         type = "PubChemFingerprint")

unique_structuresDis_long <- data.table(unique_structuresDis$fingerDisMat) %>% mutate(compound=row.names(unique_structuresDis$fingerDisMat)) %>% pivot_longer(cols = compound_1:compound_483, values_to= "compDis",names_to="compound_2")

comp_1_info <- unique_structures %>% rename(compound_1=compound,FT_ID_1 = FT_ID,smiles_1=smiles,inchikey_1=inchikey) %>% select("compound_1","FT_ID_1","smiles_1","inchikey_1")

comp_2_info <- unique_structures %>% rename(compound_2=compound,FT_ID_2 = FT_ID,smiles_2=smiles,inchikey_2=inchikey) %>% select("compound_2","FT_ID_2","smiles_2","inchikey_2")

unique_structuresDis_long_info <- left_join(
    left_join(data.frame(unique_structuresDis_long),comp_1_info,join_by(compound==compound_1), copy = T),
    comp_2_info,join_by(compound_2==compound_2), copy = T)
  
#For instances where a single Feature ID has differnt Smiles predictions - i.e. one from GNPS and one from Sirius - calculate similarity as the average of all predictions. 
#Note this is now a compound similarity
compDis_average <- unique_structuresDis_long_info %>% group_by(FT_ID_1,FT_ID_2) %>% summarise(compSim_mean = 1-mean(compDis))

pairwise_comps_df_all_camera_cosine_structure <- left_join(pairwise_comps_df_all_camera_cosine,compDis_average,join_by(FT_ID_1==FT_ID_1,FT_ID_2==FT_ID_2))

dim(pairwise_comps_df_all_camera_cosine_structure)

#Check that all of the comparisons get incorporated
length(which(!is.na(pairwise_comps_df_all_camera_cosine_structure$compSim_mean))) == length(compDis_average$compSim_mean)

pairwise_comps_df_all_camera_cosine_structure

```
4) Use Canopus Predictions to compare similarity based on the compDis function and "NPClassifier"
    Create an NPC_table
    Use ChemoDiv package to calculate a comp distance based on "NPClassifier"
    
```{r}

canopus_predictionsNPC_compdata <- data.frame(
compound=canopus_predictions$id, smiles = NA,inchikey=NA)


canopus_predictionsNPC_table <- data.frame(
compound=canopus_predictions$id, smiles = NA,inchikey=NA,pathway=canopus_predictions$NPC.pathway,superclass=canopus_predictions$NPC.superclass,class=canopus_predictions$NPC.class)


canopus_predictions_dist <- compDis(canopus_predictionsNPC_compdata,npcTable = canopus_predictionsNPC_table,type = "NPClassifier")



unique_NPCDis_long <- data.table(canopus_predictions_dist$npcDisMat) %>% mutate(id_1=row.names(canopus_predictions_dist$npcDisMat)) %>% pivot_longer(cols = 1:ncol(data.table(canopus_predictions_dist$npcDisMat)), values_to= "NPCDis",names_to="id_2")

dim(unique_NPCDis_long)

comp_1_info <- canopus_predictions %>% rename(id_1=id,FT_ID_1 = featureId) %>% select("id_1","FT_ID_1")
comp_2_info <- canopus_predictions %>% rename(id_2=id,FT_ID_2 = featureId) %>% select("id_2","FT_ID_2")



unique_NPCDis_long_info <- left_join(
    left_join(data.frame(unique_NPCDis_long),comp_1_info,join_by(id_1==id_1), copy = T),
    comp_2_info,join_by(id_2==id_2), copy = T)


unique_NPCDis_long_info
dim(unique_NPCDis_long_info)

NPCDis_average <- unique_NPCDis_long_info %>% group_by(FT_ID_1,FT_ID_2) %>% summarise(NPCSim_mean = 1-mean(NPCDis))

pairwise_comps_df_all_camera_cosine_structure_NPC <- left_join(pairwise_comps_df_all_camera_cosine_structure,NPCDis_average,join_by(FT_ID_1==FT_ID_1,FT_ID_2==FT_ID_2))

dim(pairwise_comps_df_all_camera_cosine_structure_NPC)

#Did all of the comparisons get incorporated
length(which(!is.na(pairwise_comps_df_all_camera_cosine_structure_NPC$NPCSim_mean))) == length(NPCDis_average$NPCSim_mean)

pairwise_comps_df_all_camera_cosine_structure_NPC


```
5) Read in Sirius substructure predictions, similar to the aproach taken by Qemistree 
      #Take Sirius predictions 
```{r}

sirius_pred_info_all  <- data.frame(id=sirius_formulas$id,FT_ID=sirius_formulas$featureId)

sirius_pred_info <- left_join(sirius_pred_info_all,sirius_pred,by="id") 

sirius_pred_info <- sirius_pred_info[!is.na(sirius_pred_info$ClassyFire.Organic.compounds),]
sirius_pred_info


sirius_frag <- data.frame(sirius_pred_info %>% select(ClassyFire.Organic.compounds:MACCS.162))

#here w use the dist function based on "euclidean" which is what is done in the qemistree paper. Should be "Jaccard" if fragment probababilities are not used.
sirius_pred_dist <- dist(data.frame(sirius_frag),diag = T,method="euclidean",upper = T)
sirius_pred_dist_dt <- as.matrix(sirius_pred_dist)
sirius_pred_dist_dt_long <- melt(sirius_pred_dist_dt, varnames = c("row", "col"))

sirius_pred_dist_dt_long$row <- as.character(sirius_pred_dist_dt_long$row)
sirius_pred_dist_dt_long$col <- as.character(sirius_pred_dist_dt_long$col)

matrix_info <- data.frame(rows = row.names(sirius_frag),id=sirius_pred_info$id, FT_ID = sirius_pred_info$FT_ID)

matrix_info_1 <- matrix_info %>% rename(rows_1=rows,FT_ID_1=FT_ID) %>% select(rows_1,FT_ID_1)
matrix_info_2 <- matrix_info %>% rename(rows_2=rows,FT_ID_2=FT_ID) %>% select(rows_2,FT_ID_2)

unique_FragDis_long_info <- left_join(
    left_join(sirius_pred_dist_dt_long,matrix_info_1,join_by(row==rows_1), copy = T),
    matrix_info_2,join_by(col==rows_2), copy = T)

unique_FragDis_long_info$FragSim <- 1-unique_FragDis_long_info$value/max(unique_FragDis_long_info$value)

FragSim_average <- unique_FragDis_long_info %>% select(FT_ID_1,FT_ID_2,FragSim) %>% group_by(FT_ID_1,FT_ID_2) %>% summarise(FragSim_mean = mean(FragSim))

pairwise_comps_df_all_camera_cosine_structure_NPC_frag <- left_join(pairwise_comps_df_all_camera_cosine_structure_NPC,FragSim_average,join_by(FT_ID_1==FT_ID_1,FT_ID_2==FT_ID_2))

dim(pairwise_comps_df_all_camera_cosine_structure_NPC_frag)

length(which(!is.na(pairwise_comps_df_all_camera_cosine_structure_NPC_frag$FragSim_mean))) == length(FragSim_average$FragSim_mean)

pairwise_comps_df_all_camera_cosine_structure_NPC_frag




```

```{R}

#2020 cosine comparisons 
table(is.na(pairwise_comps_df_all_camera_cosine_structure_NPC_frag$Cosine))

#77284 compound comparisons - similarity based on predicted structures
table(is.na(pairwise_comps_df_all_camera_cosine_structure_NPC_frag$compSim_mean))

#106929 compound comparisons - similarity based on canopus predicted classes
table(is.na(pairwise_comps_df_all_camera_cosine_structure_NPC_frag$NPCSim_mean))

#106276 compound comparisons - similarity based on predicted substructures
table(is.na(pairwise_comps_df_all_camera_cosine_structure_NPC_frag$FragSim_mean))

pairwise_comps_df_all_camera_cosine_structure_NPC_frag_summary <- pairwise_comps_df_all_camera_cosine_structure_NPC_frag %>% rowwise() %>% mutate(Compound_Similarity = mean(c(Camera_Similarity, Cosine, compSim_mean,NPCSim_mean,FragSim_mean),na.rm = T),Cosine_Camera=mean(c(Camera_Similarity, Cosine),na.rm = T))

pairwise_comps_df_all_camera_cosine_structure_NPC_frag_summary
```

Build a tree based on each of the following data

```{r}

#Choose one of the following:
vl_from <- "Cosine_Camera" # GNPS Data Only + Camera Groupings
vl_from <- "Compound_Similarity" #Average of all of the above
#** these do not work now because NAs are causing problems.
#
#vl_from <- "compSim_mean" # Based only on smiles from GNPS and Sirius hits
#vl_from <- "NPCSim_mean" # Based on Sirius Path predictions
#vl_from <- "FragSim_mean" # Based on Sirius Fragment predictions

Cmp_Sim_Matrix <- data.frame(pairwise_comps_df_all_camera_cosine_structure_NPC_frag_summary  %>% pivot_wider(id_cols = "FT_ID_1",names_from = "FT_ID_2",values_from = vl_from,values_fill = 0))
row.names(Cmp_Sim_Matrix) <- Cmp_Sim_Matrix$FT_ID_1

feature_order <- names(Cmp_Sim_Matrix)[-1][order(names(Cmp_Sim_Matrix)[-1])]

Cmp_Sim_Matrix <- data.frame(Cmp_Sim_Matrix[feature_order,feature_order])


write.table(Cmp_Sim_Matrix,here("results",paste0("custom_qemistree_compoundxcompound_distance_matrix_",vl_from,".tsv")),sep="\t")

Cmp_Sim_Matrix[is.na(Cmp_Sim_Matrix)] <- 0
Cmp_Sim_Matrix


```
Use pvclust to create Tree
```{r}
source(here("code","as.phylo.pvclust.R"))

#This builds Tree of compounds vs compounds

#Species_TIC_norm <- decostand(Species_TIC, method = "total", MARGIN = 2) 
Species_TIC_norm <- Cmp_Sim_Matrix
species_dist <- as.dist(1-cor(Species_TIC_norm, method="pearson"))				# Distance matrix
species_single <- hclust(species_dist, method="single")					# Single linkage clustering
species_ward <- hclust(species_dist, method="ward.D")					# Ward clustering
species_complete <- hclust(species_dist, method="complete")				# Complete linkage clustering
species_centroid <- hclust(species_dist, method="centroid")				# Centroid clustering
species_median <- hclust(species_dist, method="median")				        # Median clustering
#Comparison between the distance matrix and binary matrices representing partitions 
coph1 <- cophenetic(species_single)							# Compute Patristic distances		
coph2 <- cophenetic(species_ward)
coph3 <- cophenetic(species_complete)
coph4 <- cophenetic(species_centroid)
coph5 <- cophenetic(species_median)
a <-cor(coph1, species_dist)								#Cophenetic correlations
b <-cor(coph2, species_dist)
c <-cor(coph3, species_dist)
d <-cor(coph4, species_dist)
e <-cor(coph5, species_dist)
method <- c("single","ward","complete", "centroid", "median")
mhc<- method[which.max(c(a,b,c,d,e))]
print(paste("Building tree with", mhc, "clustering"))
result_species <- pvclust(Species_TIC_norm, method.hclust=mhc, method.dist="correlation", use.cor="pairwise.complete.obs", nboot=1000,parallel=T)

#Write tree as .tre file to be edited in Figtree
chemocode_tree<-as.phylo.pvclust(result_species)

write.tree(chemocode_tree,file=here("results",paste0("custom_qemistree_compoundxcompound_tree_summary_",vl_from,".tre")))
```

Create CSCS matricies using q2-CSCS 

#Note 2024 Feb 27 below this line is not working. Dale will update at some point
```{R}


system(paste(
  here("code","calculate_CSCS_matrices.sh"), # file path for bash script
  here("results"), #path for working directory
  here("data","quantification_table_V2_2023_10_10.tsv"), #path to compound quantification table 
  here("data","quantification_table_PA_V2_2023_10_10.tsv"))) #name of of PA table 

```
Build a tree out of the Q2-CSCS matricies created above
```{r}
source(here("code","as.phylo.pvclust.R"))

cscs_matrices <- list.files(here("results","gnps_output","custom_qemistree"),pattern ="cscs_distance_matrix*")

for (file in cscs_matrices) {
CSCS <- read.csv(here("results",file),sep="\t",row.names = 1)

Species_TIC_norm <- CSCS
#Species_TIC_norm <- decostand(Species_TIC, method = "total", MARGIN = 2) 
species_dist <- as.dist(1-cor(Species_TIC_norm, method="pearson"))				# Distance matrix
species_single <- hclust(species_dist, method="single")					# Single linkage clustering
species_ward <- hclust(species_dist, method="ward.D")					# Ward clustering
species_complete <- hclust(species_dist, method="complete")				# Complete linkage clustering
species_centroid <- hclust(species_dist, method="centroid")				# Centroid clustering
species_median <- hclust(species_dist, method="median")				        # Median clustering
#Comparison between the distance matrix and binary matrices representing partitions 
coph1 <- cophenetic(species_single)							# Compute Patristic distances		
coph2 <- cophenetic(species_ward)
coph3 <- cophenetic(species_complete)
coph4 <- cophenetic(species_centroid)
coph5 <- cophenetic(species_median)
a <-cor(coph1, species_dist)								#Cophenetic correlations
b <-cor(coph2, species_dist)
c <-cor(coph3, species_dist)
d <-cor(coph4, species_dist)
e <-cor(coph5, species_dist)
method <- c("single","ward","complete", "centroid", "median")
mhc<- method[which.max(c(a,b,c,d,e))]
result_species <- pvclust(Species_TIC_norm, method.hclust=mhc, method.dist="correlation", use.cor="pairwise.complete.obs", nboot=1000,parallel=T)



chemocode_tree <-as.phylo.pvclust(result_species)
write.tree(chemocode_tree,file=here("results","gnps_output","custom_qemistree_CSCS",paste0(gsub(".tsv",".tre",file))))
}

```

